Drug sensitivity prediction with the normal inverse Gaussian prior informed by external data

### 2019-11-27 docs/manuscript1.Rnw
added EMSE split for large and small betas to simulations_gdsc.R

	\section{Simulations}\ref{sec:simulations}
	\subsection{Setup}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

@@

	We estimate four models: (i) an NIG model that does include external 
	covariates in the estimation called NIG1 in the following, (ii) the NIG model
	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
	lasso and (iv) ridge models.
	Exclusion of the external covariates as in the NIG1 model
	amounts to direct estimation of one common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines. We repeat the whole
	simulation 50 times.

@@

	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
	measures for simulation version 1. The table includes as a reference 
	the null model, which
	is an empty model that only fits the means of the $\y_{d}$.
	The median evidence lower bounds (ELBO) approximates the marginal
	likelihoods of models NIG1 and NIG2 on both training and test data, while 
	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
	performance. Focussing on MSE, we see that the NIG2 model, that includes
	the external covariates, performs best, both in terms of estimation and 
	prediction. Model fit is best for the NIG1 model that does not include the
	external covariates, according to the ELBO, calculated on the test data.
	The lower MSEs confirm that the model that
	includes the external data (NIG2) learns the underlying structure 
	in the data better than model NIG1 that does not include the external 
	data.
<<simulations_gdsc_tab1>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 1, with the best performing model in bold.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

@@

<<simulations_gdsc_tab2>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 2, with the best performing model in bold.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

@@

<<simulations_gdsc_tab3>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 3, with the best performing model in bold.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

	\section{GDSC data}\label{sec:gdsc}
	\subsection{Primary data}
	The GDSC project's \cite[]{yang_genomics_2013} aim is ``to improve cancer 
	treatments by discovering therapeutic biomarkers that can be used to identify 
	patients most likely to respond to anticancer drug". Part of the project is to
	screen $>1000$ human cancer cell lines for drug sensitivities. The cell lines
	have been genetically characterised and several drug sensitivity measures are
	recorded.
	The data is freely available from \cite{garnett_systematic_2012} and consist 
	of: (i) the sensitivity measures of the cell lines 
	to the drugs, (ii) annotation of the 
	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
	copy numbers, methylation profiles, and gene expression). We will attempt to 
	predict drug sensitivities of the cell lines, as quantified by half maximal
	inhibitory concentration (IC50), using the
	gene expression data. Other choices of sensitivity measures than IC50 are 
	possible, but a discussion on the pros and cons of different sensitivity
	measures is beyond the aim of this paper. A comprehensive penalized regression
	approach such as in 
	{\cite{mai_composite_2019} uses most of the available molecular markers 
	instead of just gene expression (68 mutations, 426 copy numbers, 
	and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
	current version of our software is not able to handle such problem sizes,
	due to the repeated calculation of posterior parameters in the empirical
	Bayes iterations. Currently, problems with number of outcomes in the hundreds
	and number of features and samples in the thousands are computationally 
	feasible.

	We average repeated measures over cell line-drug combinations and model the
	logarithm of the IC50 values. In the following, IC50 refers 
	to the log-transformed values. 
	After iteratively removing the cell line or drug with the highest proportion 
	of missings until no more missing values remain, we end up with IC50 
	estimates for 482 cell lines on 168 drugs.

@@

	\subsection{Results}
	As a preliminary step, we select
	features according to three strategies. The first strategy selects for each
	drug separately 50 features that are not in the pathway of the drug with
	maximum variance and 50 features that are in the pathway of the drug
	with maximum variance. If less than 50 features in the pathway are available,
	we select all available features. This strategy ensures
	that the features are somewhat balanced with respect to the external feature
	covariate, i.e., pathway membership. The second strategy selects the 
	100 features with the largest variance, irrespective of pathway membership,
	while the last stategy selects 1000 features with the largest variance,
	again irrespective of pathway membership. These three strategies lead to
	three different analyses described below.

	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: four models like in (\ref{eq:prior}) but with different external 
	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 does 
	not use the external covariates and estimates
	one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
	fixed.

	In addition, we estimate regular lasso and ridge models, and a
	bSEM model \cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The bSEM model 
	is similar to the NIG model in that it
	draws the $\beta_{jd}$ from a conditionally normal distribution, but
	fixes $\tau_d^2=1$ and draws
	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
	from the inverse Gaussian distribution. Additionally,
	the bSEM model can
	only incorporate one binary external covariate that divides the features into
	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1
	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix}
	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both
	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
	estimated in a similar fashion as in Section \ref{sec:estimation}.
	The null model that fits the means of the $\y_d$ is included as a reference.

@@

tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="elbot", ], 2, median) -
               apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
                   "$\\text{ELBO}_{\\text{train}} 
                   - \\text{ELBO}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$")  
@
  From the $\bm{\alpha}$ estimates in Tables \ref{tab:analysis_gdsc_est1_feat} 
  and \ref{tab:analysis_gdsc_est1_feat} we see that there
  seems to be an effect of the
  external variables. Models NIG2 and NIG3 estimate a negative effect for the
  pathway external covariate. This translates to a negative additive effect on
  the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are found
  in the target pathway of a drug are more predictive of IC50 than genes that
  are not in the target pathway, according to expectation. Furthermore the NIG2
  model estimates positive effects for the experimental and developmental drugs,
  compared to the approved drugs. This translates to smaller expected 
  prior precisions for the regression parameters for approved drugs, i.e.,
  approved drugs are easier to predict than experimental and developmental 
  drugs. On the other hand, the NIG4 model, that only includes external
  drug covariates, estimates a negative effect for the developmental drugs.

  If we look at the NIG1 model, that includes all external 
  covariates, we see that approved drugs are easier to predict, due to small
  prior precision, followed by drugs that are in clinical development,
  as we expected. The cytotoxic drugs are the hardest to predict, 
  with the largest prior precision (smallest prior variance), followed by the 
  targeted drugs and the unknown drugs. This is again according to expectation.
  For the unkown drugs, we are unable to confirm the small precision. Genes 
  that are in the target pathway have 
  smaller prior precisions, according to expectation, which confirms that these
  genes are more important for prediction of drug sensitivity. The estimates 
  from
  the bSEM model are also according to expectation: genes that are in the target
  pathway have a much smaller precision than genes that are not in the target
  pathway. The 
  effects of the pathway indicator seem to be much larger in the bSEM model. 
<<analysis_gdsc_est1_feat>>=
options(knitr.kable.NA="", digits=3)
kable(est1,
      col.names=colnames(est1), align="r",
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 1.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<analysis_gdsc_est1_drug>>=
kable(est2,
      col.names=colnames(est2), align="r",
      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 1.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  The median prediction MSEs and the median ELBOs, calculated on the test data
  are displayed in Table \ref{tab:results_data}. In terms of 
  is lowest in the NIG models 
  (\Sexpr{round(rnorm(1))} and \Sexpr{rnorm(1)}, in the
  NIG1 and NIG2 models, respectively). The bSEM model is not able to 
  improve on the null model with an median prediction MSE of 
  \Sexpr{round(rnorm(1))} compared to the null model's MSE of 
  \Sexpr{round(rnorm(1))}.
<<analysis_gdsc_tab1>>=
kable(tab,
      col.names=colnames(tab), align="r",
      caption="Performance measures from GDSC data for feature selection strategy 1.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  In addition to the empirical Bayes estimates and the prediction MSE, we 
  investigated the stability of the estimates by ranking the posterior means
  $\E_{Q}(\bbeta_d)$ and calculating the Spearman's rank
  correlation over the 50 data splits. They are depicted in Figure 
  . In both NIG models the correlations
  between splits are higher than in the bSEM model; an indication that 
  estimation in the NIG model is more stable than in the bSEM model.

<<analysis_gdsc_fit2>>=
fit <- read.table("results/analysis_gdsc_fit2.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

res <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
colnames(est1) <- c("intercept", "not in pathway", "in pathway")
est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
colnames(est2) <- c("intercept", "experimental", "development", "approved",
                    "unkown", "targeted", "cytotoxic")

tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="elbot", ], 2, median) -
               apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
                   "$\\text{ELBO}_{\\text{train}} 
                   - \\text{ELBO}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$")  
@

<<analysis_gdsc_est2_feat>>=
options(knitr.kable.NA="", digits=3)
kable(est1,
      col.names=colnames(est1), align="r",
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 2.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<analysis_gdsc_est2_drug>>=
kable(est2,
      col.names=colnames(est2), align="r",
      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 2.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<analysis_gdsc_tab2>>=
kable(tab,
      col.names=colnames(tab), align="r",
      caption="Performance measures from GDSC data for feature selection strategy 2.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<analysis_gdsc_fit3>>=
fit <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

res <- read.table("results/analysis_gdsc_res3.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

est1 <- cbind(fit[, 1], c(NA, 0, 0, NA), fit[, 2])
colnames(est1) <- c("intercept", "not in pathway", "in pathway")
est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0), fit[, c(7:6)], c(NA, 0, NA, 0))
colnames(est2) <- c("intercept", "experimental", "development", "approved",
                    "unkown", "targeted", "cytotoxic")

tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="elbot", ], 2, median) -
               apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
                   "$\\text{ELBO}_{\\text{train}} 
                   - \\text{ELBO}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$")  
@

>>>>

	\section{Simulations}\label{sec:simulations}
	\subsection{Setup}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

@@

	We estimate four models: (i) the NIG model that only includes an intercept in
	the external covariates, called NIG$_{\text{f}}^-$, NIG$_{\text{d}}^-$, or
	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
	estimated,
	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
	the all external covariates, called 
	NIG$_{\text{f}}$, NIG$_{\text{d}}$, or
	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
	estimated, and regular (iii) lasso and (iv) ridge models.
	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
	NIG$_{\text{d}}^-$, and NIG$_{\text{f}+\text{d}}^-$ models
	amounts to direct estimation of one common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in the 
	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models. 
	All models are fit on 241 randomly selected cell lines, while performance
	measures are estimated from the remaining 241 cell lines. We repeat the whole
	simulation 50 times.

@@

	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
	measures for simulation version 1. The table includes as a reference 
	the null model, which
	is an empty model that only fits the means of the $\y_{d}$.
	The median evidence lower bounds (ELBO) approximates the marginal
	likelihoods of models NIG1 and NIG2 on both training and test data, while 
	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
	performance. Focussing on MSE, we see that the NIG2 model, that includes
	the external covariates, performs best, both in terms of estimation and 
	prediction. Model fit is best for the NIG1 model that does not include the
	external covariates, according to the ELBO, calculated on the test data.
	The lower MSEs confirm that the model that
	includes the external data (NIG2) learns the underlying structure 
	in the data better than model NIG1 that does not include the external 
	data.
<<simulations_gdsc_tab1>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 1 (best performing model in bold).", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

@@

<<simulations_gdsc_tab2>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 2 (best performing model in bold).", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

@@

<<simulations_gdsc_tab3>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 3 (best performing model in bold).", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

	\section{GDSC data}\label{sec:gdsc}
	\subsection{Primary data}
	The GDSC project's \cite[]{yang_genomics_2013} aim is ``to improve cancer 
	treatments by discovering therapeutic biomarkers that can be used to identify 
	patients most likely to respond to anticancer drug". Part of the project is to
	screen $>1000$ human cancer cell lines for drug sensitivities. The cell lines
	have been genetically characterised and several drug sensitivity measures are
	recorded.
	The data is freely available from \cite{garnett_systematic_2012} and consist 
	of: (i) the sensitivity measures of the cell lines 
	to the drugs, (ii) annotation of the 
	screened compounds, and (iii) the cell lines' genomic profile (mutations, 
	copy numbers, methylation profiles, and gene expression). We will attempt to 
	predict drug sensitivities of the cell lines, as quantified by half maximal
	inhibitory concentration (IC50), using the
	gene expression data. Other choices of sensitivity measures than IC50 are 
	possible, but a discussion on the pros and cons of different sensitivity
	measures is beyond the aim of this paper. We have used the version of the 
	data that is presented in \cite{iorio_landscape_2016}.
	We average repeated measures over cell line-drug combinations and model the
	logarithm of the IC50 values. In the following, IC50 refers 
	to the log-transformed values. 
	After iteratively removing the cell line or drug with the highest proportion 
	of missings until no more missing values remain, we end up with IC50 
	estimates for 482 cell lines on 168 drugs.

	A comprehensive penalized regression
	approach such as in 
	{\cite{mai_composite_2019} uses most of the available molecular markers 
	instead of just gene expression (68 mutations, 426 copy numbers, 
	and 2602 gene expression levels for 498 cell lines and 97 drugs). The 
	current version of our software is not able to handle such problem sizes,
	due to the repeated calculation of posterior parameters in the empirical
	Bayes iterations. Currently, problems with number of outcomes in the hundreds
	and number of features and samples in the thousands are computationally 
	feasible.

@@

	\subsection{Results}
	A preliminary step selects, for each
	drug separately, 50 features that are not in the pathway of the drug with
	maximum variance over the cell lines and 50 features that are in the pathway 
	of the drug with maximum variance over the cell lines. If less than 50 
	features in the pathway are available,
	we select all available features in the pathway. This strategy ensures
	that the features are somewhat balanced with respect to the external feature
	covariate, i.e., pathway membership.

	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: four models like in (\ref{eq:prior}) but with different external 
	variables, named NIG1 and NIG2 in the following. NIG1 does 
	not use the external covariates and estimates
	one overall $\phi$ and one overall $\chi$, while NIG2 uses the external 
	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
	fixed.

	In addition, we estimate regular lasso and ridge models, and a
	bSEM model \cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The bSEM model 
	is similar to the NIG model in that it
	draws the $\beta_{jd}$ from a conditionally normal distribution, but
	fixes $\tau_d^2=1$ and draws
	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
	$\gamma_{jd}^2 \sim \mathcal{IG} (\phi_{jd}, \lambda_{jd})$. Additionally,
	the bSEM model can
	only incorporate one binary external covariate that divides the features into
	two groups, i.e., $\bm{\alpha}_{\text{feat}} = \begin{bmatrix} \alpha_0 &
	\alpha_1
	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix}
	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both
	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
	estimated in a similar fashion as in Section \ref{sec:estimation}.
	The null model that fits the means of the $\y_d$ is included as a reference.

@@

tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="elbot", ], 2, median) -
               apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median))
tabm <- tab
tabm <- round(tabm, 3)
tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tabm[which(tabm[, 1]==max(tabm[, 1], na.rm=TRUE)), 1], "}")
tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tabm[which(tabm[, 2]==min(tabm[, 2], na.rm=TRUE)), 2], "}")
tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tabm[which(tabm[, 3]==min(tabm[, 3], na.rm=TRUE)), 3], "}")
colnames(tabm) <- c("$\\text{ELBO}_{\\text{test}}$",
                   "$\\text{ELBO}_{\\text{train}} 
                   - \\text{ELBO}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$")

library(glmnet)
kurtosis <- function(x) {
  n <- length(x) 
  kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) - 
    3*(n - 1)^2/((n - 2)*(n - 3))
  replace(kurt, !is.finite(kurt), NA)
}
load("results/analysis_gdsc_fit1.Rdata")
best2.semnig <- fit2.semnig$vb$mu
best1.lasso <- lapply(fit1.lasso, function(s) {
  unname(coef(s, "lambda.min")[-1, ])})
best1.ridge <- lapply(fit1.ridge, function(s) {
  unname(coef(s, "lambda.min")[-1, ])})
best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE), 
           mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
           mean(sapply(best1.ridge, kurtosis), na.rm=TRUE), 
           mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
ranges <- rbind(range(unlist(lapply(1:length(
  fit2.semnig$eb$mpriord), function(d) {
    fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
  fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
    (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
@
  The non-zero $\bm{\alpha}$ estimates in Tables 
  \ref{tab:analysis_gdsc_est1_feat} and \ref{tab:analysis_gdsc_est1_drug} show 
  that there is an effect of the
  external covariates. Models NIG2, NIG3, and bSEM estimate a negative effect 
  for
  the external pathway covariate. This translates to a negative additive effect 
  on the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are 
  found
  in the target pathway of a drug are more predictive of IC50 than genes that
  are not in the target pathway, according to expectation. Furthermore the NIG2
  model estimates positive effects for the experimental and developmental drugs,
  compared to the approved drugs. This translates to smaller expected 
  prior precisions for the regression parameters for approved drugs, i.e.,
  approved drugs are easier to predict than experimental and developmental 
  drugs. On the other hand, the NIG4 model, that only includes external
  drug covariates, estimates a negative effect for the developmental drugs.
  The NIG2 model estimates the smallest prior $\bm{\beta}_d$ precision for the
  targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
  are assigned the largest prior $\bm{\beta}_d$ precision, according
  to expectation. Somewhat surprisingly, the NIG4 model estimates the smallest
  prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
  targeted and unkown drugs. We conjecture that, due to absence of a local
  feature-specific variance component, this model is not flexible 
  enough to correctly capture the full covariance structure of the 
  $\bm{\beta}_d$ and therefore incorreclty estimates $\bm{\alpha}$.
  To see the total effect of the external covariates on the $\bm{\beta}_d$ 
  prior, we calculated the range of the expected prior variances 
  (up to a multiplicative error variance) as 
  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[1, ]}\]$ and
  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[2, ]}\]$ for
  NIG2 and bSEM, respectively. This confirms that the estimated prior 
  $\bm{\beta}_d$ variance components are in the same order of magnitude and that
  the external covariates have a detectable effect on the prior.
<<analysis_gdsc_est1_feat>>=
options(knitr.kable.NA="", digits=3)
kable(est1,
      col.names=colnames(est1), align="r",
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<analysis_gdsc_est1_drug>>=
kable(est2,
      col.names=colnames(est2), align="r",
      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@
  The median PMSEs and ELBOs, calculated on the test data
  are displayed in Table \ref{tab:analysis_gdsc_tab1}. In terms of prediction 
  error, ridge outperforms the other models 
  (PMSE of \Sexpr{round(tab[6, 3], 3)}), while bSEM 
  (PMSE of \Sexpr{round(tab[7, 3], 3)}) performs worse than 
  the null model (PMSE of \Sexpr{round(tab[8, 3], 3)}). Second and third best
  performing models are NIG1 and NIG2 (PMSEs of \Sexpr{round(tab[1, 3], 3)} and
  \Sexpr{round(tab[2, 3], 3)}, respectively), both outperforming the lasso
  \Sexpr{round(tab[5, 3], 3)}. Part of the differences in performance between
  NIG, lasso, and ridge may be explained with the different levels of 
  ``sparsity'' in the solution. Although NIG and ridge do not give exactly 
  sparse solutions (none of the $\bm{\beta}_d$ coefficients are set to zero), we
  may use the ``heavy-tailedness'' of the distribution of the estimated 
  $\bm{\beta}_d$ as a proxy for sparsity. A common measure of heavy-tailedness
  is excess kurtosis, calculated as \Sexpr{round(kurts[1], 3)}, 
  \Sexpr{round(kurts[2], 3)}, and \Sexpr{round(kurts[3], 3)} for the NIG2, 
  lasso, and ridge estimates, respectively. Evidently, lasso gives the 
  ``sparsest'' solution, followed by NIG and ridge. In some sense, this
  means that the NIG uses less of the $\bm{\beta}_d$ to predict.
<<analysis_gdsc_tab1>>=
kable(tabm,
      col.names=colnames(tabm), align="r",
      caption="Performance measures calculated on GDSC data (best performing model in bold).",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

	\section{Discussion}\label{sec:discussion}
<<time>>=
load("results/analysis_gdsc_fit1.Rdata")
time2.semnig <- sum(fit2.semnig$time)
@	
	The preceding presents a novel model for drug sensitivity
	prediction from a set of high dimensional molecular features. The model allows
	for the inclusion of discrete and continuous external covariates on both the 
	drugs and features. Inclusion of the external information is through 
	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
	the normal inverse Gaussian prior model (\ref{eq:prior}). Estimation is 
	efficient and scales well with the number of features and samples.
	Estimation of the NIG model that includes both external drugs and feature
	covariates (NIG2 in Section \ref{sec:gdsc}) on the GDSC data from Section
	\ref{sec:gdsc} took \Sexpr{round(time2.semnig, 2)} seconds on a 2016 
	MacBook Pro with a 2 GHz Dual-Core Intel Core i5 processor and
	8 GB of 1867 MHz LPDDR3 memory, running macOS  10.15.1.

	Simulations in Section \ref{sec:simulations} show that the method is 
	reasonably able to recover hyperparameters. In addition, the simulations
	show that inclusion of informative external covariates is indeed beneficial to
	predictive performance. The model is put into practice on the GDSC data. 
	The results in Section
	\ref{sec:gdsc} show that the inclusion of the external covariates is 
	beneficial for drug sensitivity prediction. Prediction MSE of the NIG
	model with both both feature and drug variance components is lower
	than for the regular lasso model and although the prediction MSE is
	not lower if external covariates are included, the resulting hyperparameter 
	estimates are still informative and may guide future research efforts.
	In addition, the comparison to the bSEM model that allows for the inclusion of 
	one binary external feature covariate favours the proposed NIG model. 
	Prediction MSEs are almost uniformly lower over the drugs for the NIG model 
	compared to bSEM. Consequently, the total prediction MSE is significantly
	lower for the NIG model.
